Functional Pearl: the Proof Search Monad

### 2015-11-10 paper/main.tex
More excellent remarks from reviewer #2.

\usepackage{enumerate}
\usepackage{tabularx}
\usepackage{listings}
  \def\li{\lstinline}
  \lstset{
    language=caml,
    flexiblecolumns=false,
    showstringspaces=false,
    basicstyle=\ttfamily,
    framesep=5pt,
    numberstyle=\tiny,
    numbersep=5pt,
    escapeinside={(*}{*)},
    deletekeywords={value},
  }
\usepackage{minted}
  \newminted[ocaml]{ocaml}{mathescape,fontsize=\small}
\usepackage{mathpartir}
  % Improvements to Didier's mathpartir package: make all rule names and
  % references use textsc, and also make references to rules done using the \Rule
  % command clickable in a PDF.
  \let\TirName\textsc
  % \renewcommand{\RefTirName}[1]{\hypertarget{#1}{\TirName {#1}}}
  \renewcommand{\DefTirName}[1]{\hyperlink{#1}{\TirName {#1}}}
  % \let\DefRule\RefTirName
  \let\Rule\DefTirName
  % \let\Rule\textsc

@@

This paper attempts to present, in a tutorial-style, the design of an OCaml
library. In order to facilitate the discussion, we focus on a very constrained
logic; later (\sref{extending}), we mention how to extend the library to
cover more use-cases. The original motivation for the library was to serve as a
core building block for the type-checker of Mezzo~\cite{protzenko-phd-14}. The
nature of the core, minimal logic that we are about to present is, of course,
inspired by typical nature of type-checking problems: it features equality,
quantifiers, and positive literals; \sref{extending} mentions how to extend it
with function symbols and variance (negative positions), as is typical for
type-checking problems.

@@

For instance, in our logic of existentially-quantified conjunctions of
equalities, one may want to use the rules from \fref{proof-system}. These rules
differ from \fref{logic} in that they are algorithmic; they take an input and
return an output.

@@

    \inferrule[Inst]{
      x \in V \\ \f y \in V \\ \f y \not\in \sigma \\\\
      V, \{ \f y \mapsto x \} \circ \sigma \vdash  P \dashv \sigma'
    }{
      V \vdash P \dashv \sigma'
    }

    \inferrule[And]{
      V, \sigma \vdash P \dashv \sigma' \\\\
      V, \sigma' \vdash Q \dashv \sigma''
    }{
      V \vdash P \wedge Q \dashv \sigma''
    }

@@

let unify state v1 v2: state =                  let rec solve state formula: state =
  match P.find v1 state, P.find v2 state with     match formula with
  | Flexible, Flexible                            | Equals (v1, v2) ->
  | Flexible, Rigid ->                                unify state v1 v2
      return (P.union v1 v2 state)                | And (f1, f2) ->
  | Rigid, Flexible ->                                solve state f1 >>= fun state ->
      return (P.union v2 v1 state)                    solve state f2
  | Rigid, Rigid ->
      if P.same v1 v2 state then
        return state
      else
        nothing
  \end{ocaml}
  \caption{Solver for the simplified problem}
  \label{fig:solver}
\end{figure}

@@

The type of computations \li+'a MWriter.t+ boils down (after functor
application) to \li+(derivation list * state) option+. A computation in the monad
represents a given point in the proof; the solver is focused on a rule; has
proved a number of premises so far (the \li+derivation list+); has reached a
certain \li+state+ (threaded through the premises). The \li+option+ type
accounts for failure; in case a premise cannot be proved, the computation aborts
and becomes \li+None+.

@@

The approach advocated in the present paper works well for sequent-style
calculi; in the context of Mezzo, the logic is extended with the following extra
features:
\begin{itemize}
  \item function symbols, such as, but not limited to: ML arrow types ($\to$),
    type applications ($\mathsf{list}\ \alpha$), constructor applications
    ($\mathsf{Cons}\;\{ \mathsf{head}:\alpha;\ \mathsf{tail}:\mathsf{list}\
    \alpha\}$)
  \item positive and negative positions (also known as ``variance'' in
    type-checking lingo; this applies to function symbols, such as arrows or
    type constructors)
  \item higher-order quantification ($\forall (p: \mathsf{predicate}) \ldots$)
  \item affinity (where some hypothesis may be used at most once)
  \item framing, which bears some similarities with focusing.
\end{itemize}

@@

For the complete system (Mezzo), there are many more backtracking points. The
client code choses to backtrack on only a subset of them.

@@

@@

It is unclear how one could memoize the sub-computations; the state of the
client code (the union-find) may not change very much in the simplified logic
presented here; however, for a more evolved logic, it is likely that the client
state evolves monotonically, and makes memoizing difficult, as the result of
sub-calls depends very much on the state.

>>>>

\usepackage{enumerate}
\usepackage{tabularx}
\usepackage{listings}
  \def\li{\lstinline}
  \lstset{
    language=caml,
    flexiblecolumns=false,
    showstringspaces=false,
    basicstyle=\ttfamily,
    framesep=5pt,
    numberstyle=\tiny,
    numbersep=5pt,
    escapeinside={(*}{*)},
    deletekeywords={value},
  }
\usepackage{minted}
  \newminted[ocaml]{ocaml}{mathescape,fontsize=\small}
\usepackage{mathpartir}
  % Improvements to Didier's mathpartir package: make all rule names and
  % references use textsc, and also make references to rules done using the \Rule
  % command clickable in a PDF.
  \let\TirName\textsc
  \renewcommand{\RefTirName}[1]{\hypertarget{#1}{\TirName {#1}}}
  \renewcommand{\DefTirName}[1]{\hyperlink{#1}{\TirName {#1}}}
  \let\DefRule\RefTirName
  \let\Rule\DefTirName
  % \let\Rule\textsc

@@

This paper attempts to present, in a tutorial-style, the design of an OCaml
library. In order to facilitate the discussion, we focus on a very constrained
logic; later (\sref{extending}), we briefly discuss how to extend the library to
cover more use-cases. The original motivation for the library was to serve as a
core building block for the type-checker of Mezzo~\cite{protzenko-phd-14}. The
nature of the core, minimal logic that we are about to present is, of course,
inspired by typical nature of type-checking problems: it features equality,
quantifiers, and positive literals; \sref{extending} mentions how to extend it
with, among other things, function symbols and variance (positive/negative
positions), as is typical for type-checking problems.

@@

For instance, in our logic of existentially-quantified conjunctions of
equalities, one may want to use the rules from \fref{proof-system}. These rules
differ from \fref{logic} in that they are algorithmic; they take an input
($\vdash$) and
return an output ($\dashv$).

@@

    \inferrule[Inst]{
      x \in V \\ \f y \in V \\ \f y \not\in \sigma \\\\
      V, \{ \f y \mapsto x \} \circ \sigma \vdash  P \dashv \sigma'
    }{
      V, \sigma \vdash P \dashv \sigma'
    }

    \inferrule[And]{
      V, \sigma \vdash P \dashv \sigma' \\\\
      V, \sigma' \vdash Q \dashv \sigma''
    }{
      V, \sigma \vdash P \wedge Q \dashv \sigma''
    }

@@

let unify state v1 v2: state option =           let rec solve state formula: state =
  match P.find v1 state, P.find v2 state with     match formula with
  | Flexible, Flexible                            | Equals (v1, v2) ->
  | Flexible, Rigid ->                                unify state v1 v2
      return (P.union v1 v2 state)                | And (f1, f2) ->
  | Rigid, Flexible ->                                solve state f1 >>= fun state ->
      return (P.union v2 v1 state)                    solve state f2
  | Rigid, Rigid ->
      if P.same v1 v2 state then
        return state
      else
        nothing
  \end{ocaml}
  \caption{Solver for the simplified problem}
  \label{fig:solver}
\end{figure}

@@

The type of computations \li+'a MWriter.t+ boils down to \li+(derivation list * state) option+
after functor application. A computation in the monad
represents a given point in the proof; the solver is focused on a rule; has
proved a number of premises so far (the \li+derivation list+); has reached a
certain \li+state+ (threaded through the premises). The \li+option+ type
accounts for failure; in case a premise cannot be proved, the computation aborts
and becomes \li+None+.

@@

The approach advocated in the present paper works well for sequent-style
calculi; in the context of Mezzo, the logic is extended with the following extra
features:
\begin{itemize}
  \item function symbols, such as, but not limited to: ML arrow types ($\to$),
    type applications ($\mathsf{list}\ \alpha$) and constructor applications
    ($\mathsf{Cons}\;\{ \mathsf{head}:\alpha;\ \mathsf{tail}:\mathsf{list}\
    \alpha\}$)
  \item positive and negative positions (also known as ``variance'' in
    type-checking lingo; this applies to function symbols, such as arrows or
    type constructors)
  \item higher-order quantification ($\forall (p: \mathsf{predicate}) \ldots$)
  \item affinity (where some hypothesis may be used at most once)
  \item framing, which bears some similarities with focusing.
\end{itemize}

@@

@@

The logic does not necessarily have to be decidable for this approach to work
well; indeed, we conjecture that type-checking Mezzo programs is not
decidable~\cite[p.  167]{protzenko-phd-14}. What matters is that exploration
follows a deterministic set of rules; in Mezzo, the backtracking points are
chosen and controlled~\cite[p. 165]{protzenko-phd-14}. We, of course, explore a
fine-tuned subset of the search space.

@@

It is unclear how one could memoize the sub-computations, as they depend very
much on the current state, which is likely to change at every step.
